TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-09-20 manuscript/manuscript.Rmd
Describe new filtering in Methods
Current WIP

The coherence of phenotype-associated gene sets should be measured with respect to the reference level of networks with paradigmatically high and low coherence. To establish a reference of high coherence, we considered networks from collections of the Molecular Signatures Database (MSigDB v6.2) [@Subramanian:2005aa]. MSigDB contains sets of genes having various types of functional relationships. We found that pathways from the Kyoto Encyclopedia of Genes and Genomes (KEGG) [@Kanehisa:2000aa] had the smallest average slope corresponding to the highest level of coherence (Supplementary Table S2). Networks assembled from Gene Ontology - Cellular Component (GOCC) collection and REACTOME pathways followed, representing alternative sources of networks of high coherence. This is expected as genes expressed in the same cellular components (GOCC genes) or participating in the same metabolic pathwaays (REACTOME genes) are presumed to interact more frequently. In our study, KEGG networks were used as the primary reference of high coherence, in parallel with GOCC and REACTOME collections. Conversely, as a reference for low coherence, we randomly sampled genes from a pool of genes in a given PPI database. These random gene lists represent expected network coherence arising by chance. The slopes of KEGG and random networks represent reference levels for high and low coherence, respectively.

@@

Given the dependence of internal and external degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we create size-matched references of high and low coherence. Specifically, we select the ten KEGG pathways closest in size and ten random networks matched in size as references for high/low coherence for the corresponding phenotype. Consequently, we use the slopes of internal vs. external degree distributions created by these size-matched references to normalize the slope of the phenotype-specific network to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed removes the dependency between coherence and network size (Supplementary Figure S1) and allows for direct comparison of normalized coherences across phenotypes.

@@

Phenotype-associated gene networks were formed by mapping trait-associated genes to PPIs. Genes without PPI annotations (e.g., C5orf4, LOC100507462), antisense transcripts (e.g., CDKN2B-AS), readthrough (e.g., NPHP3-ACAD11), non-protein-coding transcripts (e.g., LINC00478) were excluded. Genes that do not have PPIs with each other but interact with other genes in the global PPI network were kept. Phenotypes with less than five genes were omitted. Additionally, phenotypes with genes having identical internal degrees (hence, no degree distribution) were filtered due to the inability to derive a fit through the internal vs. external degree distributions (see below). These precautions were set to avoid situations where a small number of genes in a network can cause instability in the results. Importantly, the number of phenotype-associated genes having PPI annotations depends on the content of a PPI database; hence, some phenotypes may be filtered when using one but not the other PPI database to build phenotype-associated networks.

>>>>

The coherence of phenotype-associated gene sets should be measured with respect to the reference level of networks with paradigmatically high and low coherence. To establish a reference of high coherence, we considered networks from collections of the Molecular Signatures Database (MSigDB v6.2) [@Subramanian:2005aa]. MSigDB contains sets of genes having various types of functional relationships. We found that pathways from the Kyoto Encyclopedia of Genes and Genomes (KEGG) [@Kanehisa:2000aa] had the smallest average slope corresponding to the highest level of coherence (Supplementary Table S2). Networks assembled from Gene Ontology - Cellular Component (GOCC) collection and REACTOME pathways followed, representing alternative sources of networks of high coherence. This is expected as genes expressed in the same cellular components (GOCC genes) or participating in the same metabolic pathways (REACTOME genes) are presumed to interact more frequently. In our study, KEGG networks were used in parallel with GOCC and REACTOME (referred hereafter as MSigDb networks) as a reference for high coherence. Conversely, as a reference for low coherence, we randomly sampled genes from a pool of genes in a given PPI database. These random gene lists represent expected network coherence arising by chance. The slopes of MSigDb and random networks represent reference levels for high and low coherence, respectively.

@@

Given the dependence of internal and external degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we create size-matched references of high and low coherence. Specifically, we select the 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections) and ten random networks matched in size for the corresponding network of phenotype-specific genes. Consequently, we use the slopes of internal vs. external degree distributions created by these size-matched references to normalize the slope of the phenotype-specific network to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed removes the dependency between coherence and network size (Supplementary Figure S1) and allows for direct comparison of normalized coherences across phenotypes.

@@

Phenotype-associated gene networks were formed by mapping trait-associated genes to PPIs. Genes without PPI annotations (e.g., C5orf4, LOC100507462), antisense transcripts (e.g., CDKN2B-AS), readthrough (e.g., NPHP3-ACAD11), non-protein-coding transcripts (e.g., LINC00478) were excluded. Genes that do not have PPIs with each other but interact with other genes in the global PPI network were kept. A non-zero interacting genes were defined as genes having at least one interaction with other phenotype-specific genes. Consequently, phenotypes with less than 10 non-zero interacting genes were omitted. Additionally, phenotypes with genes having identical internal degrees (hence, no degree distribution) were filtered due to the inability to derive a fit through the internal vs. external degree distributions (see below). These precautions were set to avoid situations where a small number of genes in a network can cause instability in the results. Importantly, the number of phenotype-associated genes having PPI annotations depends on the content of a PPI database; hence, some phenotypes may be filtered when using one but not the other PPI database to build phenotype-associated networks.
